Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-29 joint_inf.tex
Tweaks and notes.

For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Following Section 8.6 of \cite{Semple2003}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\sitePatternSet)|$ and unique ancestral state splits as $\nAncestralSplits=|\mathcal{P}(\ancestralStateSet)|$ where $\mathcal{P}$ denotes the powerset.

%EM in general aren't a different number of internal site patterns than tip site patterns?
%das: Yes, I moved the paragraph at the end to here so it's hopefully less confusing. Maybe writing out the example would help? I also changed from "pattern" to "category" which might make it a little more clear.
%EM Seems good this is an important concept so it's worth defining it more properly.

Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) &= P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i\in\ancestralStateCategory_j | \tau, t) \\
     &= P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t).
\end{align*}
This equivalence is helpful in writing the full likelihood of the joint inference problem in terms of site splits; for most likely ancestral state partitions $\fullAncestralStateCategories=\{\ancestralStateCategory_1,\ldots,\ancestralStateCategory_\nSiteSplits\}$ we reformulate the likelihood equation
\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) \\
%EM Can you think a little about z_j?
%EM Even though it's a category now, it's still a function of tau and t, right?
                             &= \prod_{i=1}^{\nCols} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

>>>>

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
%EM What are complementary characters?
For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
%EM Why nSiteRows-1 in the def of siteSplit above, but nAncestralStateRows below in the def of ancestralSplit?
%EM There are two ways to do this, if you are collapsing such that 001 and 110 are the same or not, and we should be clear which one you are doing.
The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\sitePatternSet)|$ and unique ancestral state splits as $\nAncestralSplits=|\mathcal{P}(\ancestralStateSet)|$ where $\mathcal{P}$ denotes the powerset.

%das: Maybe writing out the example would help?
%EM: Yes, I think so. e.g. the need for partitions \ancestralStateCategory may not be obvious right away.

Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
%EM Indexing \siteSplit_j$ not defined.
%EM But, more importantly, as we discussed Z_j is defined only with respect to a single tree and branch lengths. This need to be made abundantly clear. Personally, I think that using the same index j for them is confusing.
Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) &= P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i\in\ancestralStateCategory_j | \tau, t) \\
     &= P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t).
\end{align*}
This equivalence is helpful in writing the full likelihood of the joint inference problem in terms of site splits; for most likely ancestral state partitions $\fullAncestralStateCategories=\{\ancestralStateCategory_1,\ldots,\ancestralStateCategory_\nSiteSplits\}$ we reformulate the likelihood equation
\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) \\
%EM Can you think a little about z_j?
%EM Even though it's a category now, it's still a function of tau and t, right?
                             &= \prod_{i=1}^{\nCols} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.
